\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{array}
\usepackage{siunitx}

% Upright roman subscripts for descriptive labels (IEEE / IUPAP convention).
\newcommand{\ID}{I_{\mathrm{D}}}
\newcommand{\VD}{V_{\mathrm{D}}}
\newcommand{\IS}{I_{\mathrm{S}}}
\newcommand{\VT}{V_{\mathrm{T}}}
\newcommand{\VQ}{V_{\mathrm{Q}}}
\newcommand{\IQ}{I_{\mathrm{Q}}}
\newcommand{\vd}{v_{\mathrm{d}}}
\newcommand{\id}{i_{\mathrm{d}}}
\newcommand{\gd}{g_{\mathrm{d}}}
\newcommand{\rd}{r_{\mathrm{d}}}
\newcommand{\vhd}{\hat{v}_{\mathrm{d}}}

\begin{document}

\section*{Derivation of Diode Small-Signal and Square-Law Behavior}

The current--voltage characteristic of an ideal p--n junction diode is given by the Shockley equation:
\begin{equation}
\ID = \IS \left( e^{\frac{\VD}{n \VT}} - 1 \right)
\label{eq:shockley}
\end{equation}
For forward bias well above the thermal voltage ($\VD \gtrsim 4 \VT$, i.e.\ $\VD \gtrsim \SI{100}{\milli\volt}$), the unity term is negligible and
\begin{equation}
\ID \approx \IS\, e^{\frac{\VD}{n \VT}}
\label{eq:exp-approx}
\end{equation}
where $\IS$ is the reverse saturation current, $n$ is the ideality factor, and $\VT = kT/q \approx \SI{25.9}{\milli\volt}$ at $T = \SI{300}{\kelvin}$.

Decompose the applied voltage into a DC bias $\VQ$ and a time-varying perturbation $\vd(t)$:
\begin{equation}
\VD = \VQ + \vd.
\end{equation}
Substituting and factoring,
\begin{equation}
\ID = \underbrace{\IS\, e^{\frac{\VQ}{n \VT}}}_{\displaystyle \IQ}\, e^{\frac{\vd}{n \VT}} = \IQ\, e^{\frac{\vd}{n \VT}}.
\end{equation}
The Maclaurin series of the exponential converges for all $\vd$; the practical question is how many terms are needed for a given accuracy. Expanding,
\begin{equation}
\ID = \IQ \left[ 1 + \frac{\vd}{n \VT} + \frac{1}{2!}\left(\frac{\vd}{n \VT}\right)^{\!2} + \frac{1}{3!}\left(\frac{\vd}{n \VT}\right)^{\!3} + \cdots \right].
\end{equation}
The AC component is $\id \equiv \ID - \IQ$. Defining the small-signal conductance at the Q-point as $\gd \equiv \IQ/(n \VT) = 1/\rd$, this becomes
\begin{equation}
\id = \gd \vd + \frac{\gd}{2\, n \VT}\, \vd^{\,2} + \frac{\gd}{6\,(n \VT)^2}\, \vd^{\,3} + \cdots
\label{eq:power-series}
\end{equation}

\subsection*{Distortion and regime boundaries}

For a sinusoidal drive $\vd = \vhd \cos\omega t$, each term in \eqref{eq:power-series} generates harmonics of $\omega$. The fractional second- and third-harmonic distortions relative to the fundamental are
\begin{equation}
\mathrm{HD2} \approx \frac{\vhd}{4\, n \VT}, \qquad
\mathrm{HD3} \approx \frac{\vhd^{\,2}}{24\,(n \VT)^2}.
\end{equation}
These quantitative criteria, rather than $\VT$ itself, set the boundaries between operating regimes:
\begin{itemize}
    \item HD2 $<$ \SI{1}{\percent} requires $\vhd \lesssim 0.04\, n \VT \approx \SI{1}{\milli\volt}$ --- the linear regime.
    \item HD2 $<$ \SI{25}{\percent} (HD3 $\lesssim \SI{4}{\percent}$) requires $\vhd \lesssim n \VT \approx \SI{26}{\milli\volt}$ --- the regime in which the second-order term is the dominant nonlinearity but the series is usefully truncated at quadratic order.
\end{itemize}

\emph{Terminology note.} The diode is not a true square-law device: the linear conductance $\gd$ is always present and dominates for $\vhd \ll n \VT$. ``Square-law mixing'' refers to the fact that the $\vd^{\,2}$ coefficient is the largest nonlinearity producing sum- and difference-frequency products $\omega_1 \pm \omega_2$ in a two-tone drive, not to the I--V being purely quadratic.

\subsection*{Intermediate exponential regime}

For drive amplitudes that exceed the quadratic regime but do not yet swing the diode into reverse bias ($n \VT \ll \vhd \lesssim \VQ - 4 \VT$), the power series in \eqref{eq:power-series} converges too slowly to be useful, yet the exponential approximation \eqref{eq:exp-approx} still holds throughout the cycle. The natural representation is the closed-form expansion of the full exponential.

Substituting $\vd = \vhd \cos\omega t$ into \eqref{eq:exp-approx} and letting $x \equiv \vhd / (n \VT)$,
\begin{equation}
\ID(t) = \IQ\, e^{\,x \cos\omega t}.
\end{equation}
The modified Jacobi--Anger identity expands this as a Fourier series in $\omega t$ with coefficients given by the modified Bessel functions of the first kind, $I_k$:
\begin{equation}
\ID(t) = \IQ \left[\, I_0(x) + 2 \sum_{k=1}^{\infty} I_k(x)\, \cos(k \omega t) \,\right].
\label{eq:bessel-expansion}
\end{equation}
Two features of this regime are worth noting:

\begin{enumerate}
    \item The diode generates a rich spectrum of harmonics from a single tone purely through exponential nonlinearity, with \emph{no clipping involved}. The relative weight of the $k$-th harmonic is $2 I_k(x) / I_0(x)$, which approaches unity for all $k$ as $x$ grows.
    \item The average current rises with drive amplitude: $\langle \ID \rangle = \IQ\, I_0(x)$, and $I_0(x) > 1$ for $x > 0$. This rectified DC component (and the self-bias shift it produces when the diode feeds a resistive load through a coupling capacitor) is the operating principle of large-signal envelope detectors and grid-leak / cathode-bias detectors.
\end{enumerate}

For $x \ll 1$, $I_0(x) \approx 1 + x^2/4$ and $I_1(x) \approx x/2$, recovering the quadratic-regime results. For $x \gg 1$, $I_k(x) \sim e^x / \sqrt{2\pi x}$ independent of $k$, signaling that truncating the harmonic series at any finite order is no longer meaningful --- the transition into the rectifying regime.

\subsection*{Large-signal / rectifying regime}

When $\vhd$ approaches or exceeds $\VQ$, the negative half-cycle drives $\VD$ toward zero or into reverse bias and the forward-bias approximation \eqref{eq:exp-approx} fails. The full Shockley equation \eqref{eq:shockley} must be used, and for $\VD \lesssim -4 \VT$ the diode reaches its reverse saturation floor $\ID \to -\IS$. Neither the power series \eqref{eq:power-series} nor the Bessel expansion \eqref{eq:bessel-expansion} is a useful description --- the diode acts as a half-wave rectifier with strongly asymmetric conduction.

A summary of these four operating regimes and their boundaries is provided in Table~\ref{tab:diode-regimes}.

\bigskip

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\caption{Diode operating regimes by AC amplitude $\vhd$, with $n \approx 1$ and $\VT \approx \SI{26}{\milli\volt}$.}
\label{tab:diode-regimes}
\begin{tabular}{>{\raggedright\arraybackslash}p{2.5cm} >{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{3.9cm} >{\raggedright\arraybackslash}p{3.9cm}}
\toprule
\textbf{Regime} & \textbf{Amplitude criterion} & \textbf{Dominant behavior} & \textbf{Typical application} \\
\midrule
Linear small-signal &
$\vhd \lesssim 0.04\, n \VT$ \newline (${\sim}\SI{1}{\milli\volt}$; HD2 $<$ \SI{1}{\percent}) &
$\id \approx \gd \vd$ &
Amplifier biasing, dynamic resistance $\rd = 1/\gd$. \\

Quadratic / square-law &
$\vhd \lesssim n \VT$ \newline (${\sim}\SI{26}{\milli\volt}$) &
$\id \approx \gd \vd + \dfrac{\gd}{2 n \VT} \vd^{\,2}$ &
RF mixers, microwave power detectors, frequency doublers, intermodulation analysis. \\

Intermediate exponential &
$n \VT \ll \vhd \lesssim \VQ - 4 \VT$ \newline (series fails, $\VD > 0$) &
Bessel expansion \eqref{eq:bessel-expansion}; rich harmonics without clipping; rectified DC $\langle \ID \rangle = \IQ\, I_0(x)$ &
Unclipped large-signal envelope detection, harmonic generation, mixer LO drive, oscillator amplitude limiting. \\

Large-signal / \newline rectifying &
$\vhd \gtrsim \VQ$ \newline (exponential approx.\ fails) &
Full Shockley \eqref{eq:shockley} required; asymmetric conduction; reverse-bias floor reached on negative swing &
Peak AM envelope detection, power rectification, switching, clamping. \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Numerical values for common small-signal diodes}

The framework above instantiates differently for each diode technology. $\VT = kT/q$ depends only on temperature, so the device-specific variation enters through the ideality factor $n$ (which sets the square-law boundary $n\VT$) and the quiescent forward voltage $\VQ$ (which sets the intermediate-regime upper edge $\VQ - 4\VT$). Representative values at $T = \SI{300}{\kelvin}$ and a bias current of $\IQ \approx \SI{100}{\micro\ampere}$:

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\caption{Square-law and intermediate-regime boundaries for representative small-signal diodes at $T = \SI{300}{\kelvin}$ and $\IQ \approx \SI{100}{\micro\ampere}$. Values are approximate, drawn from typical SPICE-model parameters and datasheet curves; the linear-regime (HD2 $<$ \SI{1}{\percent}) cutoff $0.04\,n\VT$ falls between \SIrange{1.1}{1.9}{\milli\volt} across all three.}
\label{tab:diode-numbers}
\begin{tabular}{l c c c c}
\toprule
\textbf{Diode (example)} & $n$ & $\VQ$ & $n\VT$ (sq-law) & $\VQ - 4\VT$ (interm.) \\
\midrule
Si PN signal (1N4148)        & 1.85 & \SI{0.55}{\volt} & \SI{48}{\milli\volt} & \SI{450}{\milli\volt} \\
Si Schottky (1N5711, BAT54)  & 1.08 & \SI{0.30}{\volt} & \SI{28}{\milli\volt} & \SI{200}{\milli\volt} \\
Ge point-contact (1N34A)     & 1.05 & \SI{0.20}{\volt} & \SI{27}{\milli\volt} & \SI{95}{\milli\volt}  \\
\bottomrule
\end{tabular}
\end{table}

Two observations follow. First, the Si PN signal diode has roughly $1.7\times$ the square-law amplitude headroom of Ge or Schottky devices, simply because its ideality factor is higher at low bias currents; a 1N4148 mixer can accept RF input up to $\sim$\SI{48}{\milli\volt} before HD2 exceeds \SI{25}{\percent}, versus $\sim$\SI{27}{\milli\volt} for a 1N34A. Second, the much lower $\VQ$ of Ge and Schottky devices means the unclipped Bessel regime is correspondingly narrow --- only $\sim$\SI{95}{\milli\volt} of headroom for Ge before hard rectification --- whereas a Si PN diode has nearly half a volt of unclipped exponential operation above the square-law boundary. This explains the historical division of labor: Ge (and later zero-bias Si Schottky) for sensitive weak-signal detection where low $\VQ$ and low $n$ matter most, and Si PN diodes for switching, clamping, and mixer service where the wider operating range is useful.

All four regime boundaries scale with absolute temperature through $\VT \propto T$: at \SI{-40}{\degreeCelsius} they shrink by $\sim$\SI{22}{\percent}; at \SI{85}{\degreeCelsius} they expand by $\sim$\SI{19}{\percent}. The $\VQ$ term in the intermediate-regime boundary is itself temperature-dependent (roughly $-2$ mV/K at constant current), partially offsetting the $\VT$ scaling on that boundary specifically.

\end{document}
